Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-09-01 doc/papers/sc2013/paper.tex
Add first draft for the pyhpc workshop paper
http://www.dlr.de/sc/en/desktopdefault.aspx/tabid-8686/14946_read-37159/

Deadline: September 6!

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{sigplanconf}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\usepackage{xcolor}
\newcommand\see{\emph{cf.\ }}

@@

\lstset{language=python,morekeywords={cimport, cdef, with},
basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray}}

\begin{document}
% update listing numbering
\renewcommand{\thelstlisting}{\arabic{lstlisting}}
%
\title{Experimenting OpenMP/Python Coupling}

\authorinfo{Serge Guelton}{{\'E}cole Normale Sup{\'e}rieure, Paris, France}{serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}{T{\'e}l{\'e}com Bretagne, Plouzan{\'e},France}{pierrick.brunet@telecom-bretagne.eu}

@@

    The Python language offers an interesting solution as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directives into a native parallel module that runs as
    fast as its naive C equivalent, while keeping the input code totally
    compatible with the Python interpreter. The approach is validated through
    the Pythran compiler.

\end{abstract}
\keywords{Python, OpenMP, C++, static compilation, scripting language.}

@@

The prohibitive performance overhead of the language implied by its interpreted
and highly dynamic nature does not prevent its usage as a language for high
performance computing. SciPy overcomes this issue through low-level routines
written in C or Fortran and encapsulated in Python \emph{native} modules.
Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, has been designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
cost.

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support, and
emphasises on the need for a backward-compatible approach.
Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
within the Python language in order to add fine-grain parallelism support to
Python in the context of a Python subset called Pythran~\cite{pythran2013}.
Static compilation of this language into C++ and the underlying runtime is
described in Section~\ref{sec:python-static}. The Pythran compiler is
benchmarked on several scientific kernels and compared to Cython in
Section~\ref{sec:validation}.

@@

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, such as the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.  The
numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
uses additional type information to generate sequential LLVM bytecode.  PyCUDA
and PyOpenCL~\cite{klockner2012} also target accelerators by mixing Python with
kernels embedded as raw strings containing accelerator code.

@@

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists of a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure. The languages targeted by OpenMP are statically
compiled. This section studies the semantical adaptation required to use the
same API for a scripting language such as Python.

While Python has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements. Pythran uses the
path taken by PLW~\cite{dongara2007}, that uses string instructions to hold such
decorations. As Python does not have anonymous block,~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Alternatively, the syntax
\texttt{if 'my annotation':} is also supported.

%Listing~\ref{lst:instruction-annotation} shows how to annotate a
%parallel loop with reduction this way.
%
%\begin{figure}
%    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
%    loop with reduction annotated with OpenMP directive.}]
%'omp parallel for reduction(+:r)'
%for auto in my_container: r += x
%  \end{lstlisting}
%\end{figure}

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. %Using the \texttt{default(none)} clause is possible
%to ensure no variable gets forgotten when building such lists.

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop in Pythran with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]):
    print i, ':', v
  \end{lstlisting}
\end{figure}

\subsection{Runtime Library}

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used to detect when OpenMP is not available. Python does
not have a preprocessor, but it is possible to catch the import exception if the
\texttt{omp} module is not found.

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if(loop_count):
  import omp
  using = sum = 0
  'omp parallel for if(using == 1)'
  for i in range(loop_count + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (loop_count *
              (loop_count + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

@@

\centering
\begin{tikzpicture}[
file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
node distance=0.6cm},
    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
distance=0.6cm}]
\node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
\node[file] (cxx) [below=of pythran] {\textbf{C++}};
\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

\draw[very thick, ->] (python) -- (pythran);
\draw[very thick,dashed, ->] (annotation) -- (python);
\draw[very thick,dashed, ->] (omp) -- (python);
\draw[very thick, ->] (pythran) -- (cxx);
\draw[very thick, ->] (cxx) -- (gxx);
\draw[very thick, ->] (boost) -- (gxx);
\draw[very thick, ->] (pythonic) -- (gxx);
\draw[very thick, ->] (gxx) -- (so);
\end{tikzpicture}
\caption{Pythran compiler workflow.}
\label{fig:pythran-compiler}
\end{figure}

@@

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measure the SLOC of the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

    \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
        \hline
        C       & 102   & 25 & 22 \\
        Python  & 30    & 676 & --\\
        Pythran & 30    & 38 &  22 \\
        Pythran+OMP    & 31    & 11 & no OpenMP support\\
        \hline
    \end{tabular}

@@

\begin{figure}[ht]
    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
    \label{fig:hyantes-speedup}
    \centering
    \includegraphics[width=.5\textwidth]{hyantes}
\end{figure}

@@

The performance of the two approaches is shown in
Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
mathematical, image-processing or linear algebra kernels. All these kernels have
been written in Cython and Python ---compatible with Pythran--- and annotated
through the mechanism of each language to exhibit parallelism, then compiled
into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.  Their
execution time when called from the Python interpreter is measured through the
\texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.  All results
are normalized against Cython sequential execution time. They show that while
handling code at a higher level than Cython, Pythran achieves comparable
results.

@@

@@

%=========================================================
\acks
%=========================================================

This work has received fundings from \textsc{silkan} and the French ANR through the
CARP project. The authors thank Adrien \textsc{Merlini} and Alan
\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Mehdi \textsc{Amini},
Fran{\,c}ois \textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable
reviews.

\bibliographystyle{abbrvnat}
\bibliography{biblio}

>>>>

% to submit via
% https://www.easychair.org/account/signin.cgi?timeout=1;conf=pyhpc2013
% max 10 pages
\documentclass[conference]{IEEEtran}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\usepackage{xcolor}
\newcommand\see{\emph{cf.\ }}

@@

\lstset{language=python,morekeywords={cimport, cdef, with},
basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray},
captionpos=b}

\begin{document}
% update listing numbering
\renewcommand{\thelstlisting}{\arabic{lstlisting}}
%
\title{Compiling Python Modules to Native Parallel Modules Using Pythran and
OpenMP Annotations}

\author{
    \IEEEauthorblockN{Serge Guelton}
    \IEEEauthorblockA{
        {\'E}cole Normale Sup{\'e}rieure, Paris, France\\
        Email: \texttt{serge.guelton@telecom-bretagne.eu}\\
    }
    \and
    \IEEEauthorblockN{Pierrick Brunet}
    \IEEEauthorblockA{
        T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France\\
        Email: \texttt{pierrick.brunet@telecom-bretagne.eu}\\
    }
    \and
    \IEEEauthorblockN{Mehdi Amini}
    \IEEEauthorblockA{
        Silkan, USA\\
        Email: \texttt{mehdi.amini@silkan.com}
    }
}

@@

    HPC users traditionally rely on low-level, compiled language such as C or
    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
    common situation to have HPC application written in a high-level language
    such as Python, calling native routines for computation-intensive tasks. To
    improve development speed and reduce maintenance costs, several solutions
    such as Cython, Numba, Parakeet or Pythran, have appeared to make it
    possible to automatically or semi-automatically turn Python functions into
    native ones.

    One of the key point to match the performance of native applications is the
    ability to write parallel applications. This paper studies the addition of
    OpenMP directives, a popular model to describe the parallelism of
    C/C++/FORTRAN applications, to Pythran, an automatic compiler from a subset
    of Python to C++. It shows that scientific Python applications annotated
    with OpenMP directives can be turned by an automatic compiler into native
    applications that run within the same order of magnitude than
    manually-written ones.

\end{abstract}

@@

It is now heading toward High Performance Computing(HPC), either as a glue
language used to bind several native libraries together, or to write whole
applications. However, the prohibitive performance overhead of the language
implied by its interpreted and highly dynamic nature should prevent its usage
for critical code sections, where native code is generally used. The SciPy
modules overcomes this issue through low-level routines written in C or Fortran
and encapsulated in Python \emph{native} modules.  Moreover, its core data
structure, the multi-dimensional array~\cite{numpyarray2011}, has been designed
so that the underlying data are available to both native modules and the Python
interpreter without conversion cost.

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} gives an
overview of existing approaches to compile Python functions or modules and shows
a critical lack of parallelism support, and emphasises on the need for a
backward-compatible approach.  Section~\ref{sec:python-static} presents the
Pythran~\cite{pythran2013} compiler, a tool that turns Python modules into C++
meta-programs, and the underlying runtime. Section~\ref{sec:python-openmp}
studies the validity of using OpenMP directives within the Python language in
order to add fine-grain parallelism support to Python in the context of Pythran.
The Pythran compiler is benchmarked on several scientific kernels and compared
to Cython in Section~\ref{sec:validation}.

@@

Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing informations, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variable to improve performance. Without type annotation, the performance
improvement is not terrific, but given enough type annotation, Cython can
generates code that runs as fast as its C equivalent, while maintaining a syntax
close to Python.

PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose an other approach
that directly mixes Python with C, using raw strings to hold the C code. PLW
also supports parallel directives that are limited to parallel for loops.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings containing accelerator code.

@@

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
compiler uses additional type information to generate sequential LLVM bytecode.
Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
Python to LLVM bytecode translation, but limits its scope to the numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
not suffer from the backward-incompatibility issues. When meeting an unsupported
construct, numba falls back to Python C-API calls, while parakeet just raises an
exception.

@@

@@

\centering
\begin{tikzpicture}[
file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
node distance=0.7cm},
    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
distance=0.7cm}]
\node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
\node[file] (omp)	   [xshift=-2em, left=of python] {\textbf{OpenMP}};
\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
\node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
\node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
\node (empty) [xshift=-1em, left=of gxx] {};
\node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic++}}};
\node[file] (annotation)	   [above=of pythonic] {\textbf{Type Info}};
\node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
\node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

\draw[very thick, ->] (python) -- (pythran);
\draw[very thick, ->] (annotation) -| (empty);
\draw[very thick,dashed, ->] (omp) -- (python);
\draw[very thick, ->] (pythran) -- (meta-cxx);
\draw[very thick, ->] (meta-cxx) -- (gxx);
\draw[very thick, ->] (boost) -| (empty);
\draw[very thick, ->] (pythonic) -- (gxx);
\draw[very thick, ->] (gxx) -- (so);
\end{tikzpicture}

\caption{Pythran compiler workflow.}
\label{fig:pythran-compiler}
\end{figure}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists of a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure. The languages targeted by OpenMP are statically
compiled. This section studies the semantical adaptation required to use the
same API for a scripting language such as Python.
Listing~\ref{lst:motivating-example} introduces a motivating example through the
computation of $\pi$, featuring a parallel reduction. The C equivalent with
OpenMP directives is given in Listing~\ref{lst:motivating-example-c} for
reference.

\begin{lstlisting}[float, label={lst:motivating-example}, caption={Motivating
example: computing $\pi$ in Python.}]
def pi(nsteps):
  sum, step = 0., 1. / nsteps
  for i in range(nsteps):
      x = (i - 0.5) * step
      sum += 4. / (1. + x**2)
  return step * sum
\end{lstlisting}

\begin{lstlisting}[float, label={lst:motivating-example-c}, caption={Motivating
example: computing $\pi$ in C with OpenMP.}]
double pi(size_t nsteps) {
  double sum = 0., step = 1. / nsteps;
  #pragma omp parallel for reduction(+:sum)
  for(size_t i = 0; i < nsteps ; ++i)
  {
      double x = (i - 0.5) * step;
      sum += 4. / (1. + x * x);
  }
  return step * sum;
}
\end{lstlisting}

While Python has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements.
PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
Python does not have anonymous block,~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Pythran uses comment to
hold OpenMP directives, and internally stores them as string instructions.
Alternatively, the syntax \texttt{if 'my annotation':} is also supported.

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. Using the \texttt{default(none)} clause is possible
to ensure no variable gets forgotten when building such lists.

The annotated motivating example is given in
Listing~\ref{lst:motivating-example-annotated-v0}. Note that unlike in C,
\texttt{x} has to be listed in a \texttt{private} clause.

\begin{lstlisting}[float, label={lst:motivating-example-annotated-v0}, caption={Motivating
example: computing $\pi$ in Python with OpenMP.}]
def pi(nsteps):
  sum, step = 0., 1. / nsteps
#omp parallel for reduction(+:sum) private(x)
  for i in range(nsteps):
      x = (i - 0.5) * step
      sum += 4. / (1. + x**2)
  return step * sum
\end{lstlisting}

\subsection{Automatic Scoping Computation}

Using function scope for all variables may lead to very long \texttt{private}
clauses. We propose an algorithm to automatically limit the scope of a variable
declaration so that it is can be automatically marked as private. The whole idea
is to compute the minimal nesting depth of a variable usage. To do so, a
\emph{nesting depth} is first associated to each expression. It lexically
corresponds to the indentation level of its instruction, to the notable
exception of the iterator declaration in a for loop, that holds a
nesting depth equals to 1 plus the loop declaration indentation level.

Tracking the nesting depth of all variable use, for instance using def-use
chains, makes it possible to build a list of nesting depths, one per variable
reference. Computing the min of this list yields the scope of the variable.

For instance, in the motivating example, the \texttt{nsteps} variable is first
defined as argument, then read twice, which gives a def-use chain of
\{DEF, USE, USE\}. The associated nesting depths is
\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
of \{D, U\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
automatically declared local to the loop, thus not needing to be listed as
\texttt{private}.

To be able to list a variable marked as local, for instance in a
\texttt{lastprivate} clause, the directive accesses are taken into account in
the def-use computations. For instance in
Listing~\ref{lst:motivating-example-annotated-v0}, the chain for \texttt{x} is
\{DEF, DEF, USE\} and the associated nested depths are \{1, 2, 2\} so its scope
is 1 and the variable is not declared local to the loop.

Thanks to this automatic scoping computation, it is possible to write shorted
OpenMP directive in a similar manner to C, as illustrated in
Listing~\ref{lst:motivating-example-annotated-v1}

\begin{lstlisting}[float, label={lst:motivating-example-annotated-v1}, caption={Motivating
example: computing $\pi$ in Python with OpenMP and automatic scoping.}]
def pi(nsteps):
  sum, step = 0., 1. / nsteps
  #omp parallel for reduction(+:sum)
  for i in range(nsteps):
    x = (i - 0.5) * step
    sum += 4. / (1. + x**2)
  return step * sum
\end{lstlisting}

\begin{lstlisting}[float, label={lst:complex-omp-for}, caption={Parallel
    loop in Pythran with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]):
    print i, ':', v
\end{lstlisting}

\subsection{OpenMP Runtime Library}

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used to detect when OpenMP is not available. Python does
not have a preprocessor, but it is possible to catch the import exception if the
\texttt{omp} module is not found. Listing~\ref{lst:sample-api-usage} showcase
such a call using an example converted from the OpenMP validation suite
presented in next section. The code starts a parallel section using the
\texttt{parallel} directive, spawning several threads. Then each started thread
increases the \texttt{nthreads} variable that has unspecified visibility, thus
\texttt{shared}. The incrementation is protected by a guard through the
\texttt{critical} directive. Then one of the thread retrieves the number of
active thread in the parallel region through the \texttt{get\_num\_threads}
function from the \texttt{omp} module. The function should always return
\texttt{True}.

\begin{lstlisting}[float,label={lst:sample-api-usage},caption={Example of
        OpenMP API usage from Python.}]
import omp
def omp_get_num_threads():
  nthreads, nthreads_lib = 0, -1
  #omp parallel
  if 1
    #omp critical
    nthreads += 1
    #omp single
    nthreads_lib = omp.get_num_threads()
  return nthreads == nthreads_lib
\end{lstlisting}

\subsubsection{Interaction With Pythran Runtime}

Pythran already uses OpenMP to parallelize some Python functions. For instance
the \texttt{sum} function from the \texttt{\_\_builtin\_\_} is implemented using
the OpenMP \texttt{reduction} clause. When it can prove that its first argument
is a \emph{pure} function, it can also call a parallel version of the
\texttt{map} function that internally spawns a parallel region. This leads to a
well known problem with OpenMP: nested parallel regions. OpenMP provides several
ways to handle this situation:

\begin{enumerate}

    \item A

    \item B

    \item C

\end{enumerate}

\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if(loop_count):
  import omp
  using = sum = 0
  'omp parallel for if(using == 1)'
  for i in range(loop_count + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (loop_count *
              (loop_count + 1)) / 2
  return known_sum == sum and num_threads == 1
\end{lstlisting}

@@

In addition to the formal validation done using the OpenMP validation suite
presented in Section~\ref{sec:python-openmp}, the performance of parallel module
generated by Pythran from Python module with explicit parallelization using
OpenMP directives has been studied on several situations: first on the
motivating example, then on a synthetic geomatic application. Python and Cython
versions are then compared and finally the effect of OpenMP directives on Pythran
generated code are illustrated in the context of the Python Benchmarks
suite~\footnote{\url{https://github.com/numfocus/python-benchmarks}}.

\subsection{Motivating Example}

To validate the approach proposed in this paper, let first consider the
motivating example. Figure~\ref{fig:motivating-example-scale} shows how the C
and Python version of the program scale when changing the number of core used.
As a reference, the original Python code runs 41 times slower than the
equivalent C code. As the program is not memory bound and exhibit trivial
parallelism, the C + OpenMP version scales linearly. The Pythran + OpenMP
version follows the same pattern.

    \caption{Comparison of the scaling of $\pi$ computation function for C and
    Python with OpenMP, depending on the number of active threads.}
    \label{fig:motivating-example-scale}

    \includegraphics[width=.5\textwidth]{pi_omp_bench.pdf}

\end{figure}

\subsection{Hyantes: a Geomatic Application}

To get closer to a realistic example, we have considered the code of a small
geomatics application, Hyantes. Starting from the C code, we successively
parallelized it with OpenMP, turned it into Python with a Pythran compatible
kernel and parallelized the Python version using the approach described in this
paper. We also measure the SLOC of the two versions.  Table~\ref{tbl:hyantes}
summarizes the results of the experimentation made on a laptop with 4 non
hyperthreaded cores using \texttt{-O2} compiler flag. It not only emphasises the
use of a high level language prototyping, but also shows that it is possible to
turn this prototype into reasonably efficient code through Pythran. It is also
possible to prototype the parallel version while remaining at the Python level.

@@

    \centering
    \begin{tabular}{|l|r|}
        \hline
        Language & SLOC  \\
        \hline
        C       & 102   \\
        Python  & 30    \\
        Pythran & 30     \\
        Pythran+OMP    & 31  \\
        \hline
    \end{tabular}

@@

    \caption{Comparison of the scaling of the hyantes application for C and
    Python with OpenMP, depending on the number of active threads.}
    \label{fig:hyantes-speedup}

    \includegraphics[width=.5\textwidth]{hyantes_omp_bench.pdf}

\end{figure}

\subsection{Comparison with Cython}

@@

The performance of the two approaches is shown in
Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
mathematical, image-processing or linear algebra kernels. All these kernels have
been written in Cython and Python ---compatible with Pythran--- and annotated
through the mechanism of each language to exhibit parallelism, then compiled
into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.  Their
execution time when called from the Python interpreter is measured through the
\texttt{timeit} module on a machine using 8 AMD Opteron 6176 SE.  All results
are normalized against Cython sequential execution time. They show that while
handling code at a higher level than Cython, Pythran achieves comparable
results.

@@

\subsection{Python Benchmarks Test Bed}

\begin{figure}[ht]
    \includegraphics[width=.5\textwidth]{growcut}
    \caption{Comparison of Various python Compilers on the Growcut benchmark}
    \label{fig:pb-growcut}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=.5\textwidth]{julia}
    \caption{Comparison of Various python Compilers on the Julia benchmark}
    \label{fig:pb-julia}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=.5\textwidth]{rosen_der}
    \caption{Comparison of Various python Compilers on the rosen\_der benchmark}
    \label{fig:pb-rosen}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=.5\textwidth]{pairwise}
    \caption{Comparison of Various python Compilers on the pairwise benchmark}
    \label{fig:pb-pairwise}
\end{figure}
growcut: $\times 2.65$
julia: $\times 3.95$
rosen\_der: $\times 1.78$
pairwise: $\times 4.27$

@@

%=========================================================
\section*{Acknowledgment}
%=========================================================

This work has received fundings from \textsc{silkan} and the French ANR through
the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois
\textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable reviews.

\bibliographystyle{IEEEtranS}
\bibliography{biblio}
